---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The data is contained within activity.zip in the working directory.  Unzip activity.csv from the zip file and load.

setwd("C:/Users/phickma/Desktop/GitHub/RepData_PeerAssessment1/activity")
setwd("C:/Users/paul_/Documents/Code/Reproducible Research/RepData_PeerAssessment1")

```{r}
unzip(zipfile = "activity.zip")
activity = read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day, ignoring the missing values in the dataset.

```{r}
totalSteps <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)

write.csv(totalSteps, "steps_per_day.csv")

hist(
    totalSteps$steps, 
    xlab = "Steps per day", 
    main = "Histogram of the total number of steps per day"
    )
```

Calculate and report the mean 
```{r}
mean(totalSteps$steps)
```

and median of the total number of steps taken per day
```{r}
median(totalSteps$steps)
```


## What is the average daily activity pattern?

A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
# interpreted the "you can ignore the missing values in the dataset" to mean this section as well
totalSteps <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)

write.csv(totalSteps, "steps_per_interval.csv")

plot(totalSteps$interval, totalSteps$steps, type = "l", xlab = "5 minute Interval", ylab = "Average Steps")
```

On average, across all the days in the dataset, the 5-minute interval that contains the maximum number of steps is:
```{r}
maxsteps <- totalSteps[totalSteps$steps==max(totalSteps$steps),]
maxsteps
```

The maximum average number of steps across all the days is **`r maxsteps$steps`** at interval **`r maxsteps$interval`**.

## Imputing missing values

There are a number of days/intervals where there are missing values. The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
# calculate the number of missing days
nomissing <- sum(is.na(activity))
nomissing
```

There are `r nomissing` missing values in the data set.

The missing days will be filled with the mean for that 5 minute interval.  A new dataset will be created will the filled in values.
```{r}

## fixsteps fills in the missing step with the average for that interval
## if the steps are not missing, then nothing happens
fixsteps <- function(steps, interval) {
    if(is.na(steps))
    {
        # totalSteps holds the mean for each interval
        # value missing so return the average for this interval
        return(totalSteps[totalSteps$interval==interval,]$steps)
    } else {
        # good
        return(steps)
    }
}

# create the new dataset with the missing values filled
activity_fixed <- activity
activity_fixed$steps <- mapply(FUN = fixsteps, activity$steps, activity$interval)

write.csv(activity_fixed, "activity_fixed.csv")
```

The histogram of the total number of steps taken each day for the new dataset is:
```{r}
totalSteps_fixed <- aggregate(steps ~ date, data = activity_fixed, sum)

write.csv(totalSteps_fixed, "steps_per_day_fixed.csv")

hist(
    totalSteps_fixed$steps, 
    xlab = "Steps per day", 
    main = "Histogram of the total number of steps per day (missing data filled)"
    )
```

For the total number of steps taken per day in the new dataset, the new mean is: 
```{r}
mean(totalSteps_fixed$steps)
```

And median is:
```{r}
median(totalSteps_fixed$steps)
```

Do these values differ from the estimates from the first part of the assignment?
The mean has not changed from the first part of the assignment.  However, the median has increased slightly to match the mean.

What is the impact of imputing missing data on the estimates of the total daily number of steps?
The missing values are for whole days rather than random intervals through the whole dataset.  As a result the imputed data inserts daily averages, which results in no change to the overall daily average.  However, the daily average is higher than some of the lower days and this has pushed up the mid point.


## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
# convert to a real date
activity_fixed$date <- as.Date(activity_fixed$date)

# create the new factor variable with weekday or weekend based on the date
activity_fixed$daytype <- factor((weekdays(activity_fixed$date) %in% c('Saturday','Sunday')), levels=c(FALSE, TRUE), labels=c('weekday', 'weekend'))
         
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(ggplot2)

# todo - datasource should be interval average

qplot(
  interval, 
  steps, 
  data = activity_fixed,
  geom = "line",
  xlab = "Interval", 
  ylab = "Number of steps", 
  main = "Comparison between weekday and weekend steps"
  ) + facet_wrap(~daytype, ncol=1)

```

Are there differences?